---
title: "Specific Analysis"
output: html_document
---
```{r setup, include=FALSE, echo =F}
knitr::opts_chunk$set(echo = TRUE)
```
<div style = "text-align: justify">
`````{r, include = F}
library(readxl)
library(gplots)
library(pheatmap)
library(VennDiagram)
`````

`````{r, eval = F}
library(readxl)
library(gplots)
library(pheatmap)
library(VennDiagram)
`````

````{r, echo =F}
## necessary data from the general analysis
meta <- read_excel("~/Documents/Uni/4.Semester/Bioinfo/GitHub/project-02-group-02/data/NCI_TPW_metadata.xls")
untreated <- readRDS("~/Documents/Uni/4.Semester/Bioinfo/GitHub/project-02-group-02/data/NCI_TPW_gep_untreated.rds")
treated <- readRDS("~/Documents/Uni/4.Semester/Bioinfo/GitHub/project-02-group-02/data/NCI_TPW_gep_treated.rds")
copynumber <- readRDS("~/Documents/Uni/4.Semester/Bioinfo/GitHub/project-02-group-02/data/CCLE_copynumber.rds")
cellline <- read_excel("~/Documents/Uni/4.Semester/Bioinfo/GitHub/project-02-group-02/data/cellline_annotation.xls")
meta = as.data.frame(meta)
celline = as.data.frame(cellline)

#scale
treated.scaled <- scale(treated)
untreated.scaled <- scale(untreated)
```````

# 2. Specific analysis
***
## Table of contents
1. [Step 1: Identification of genes as biomarker](#step1)  
    1.1 [Finding biomarker for cisplatin](#step1a)  
    1.2 [Testing biomarker genes](#significance)  
2. [Step 2: Biomarker gene expression](#upordown)  
3. [Step 3: Further evaluation of biomarker gene](#alterations)  


## Step 1: Can we identify certain genes as biomarker for cisplatin? <a name="step1"></a> 

### Step 1a: Finding Biomarker for cisplatin <a name="step1a"></a> 
#### Criterium: High fold-change values throughout the cell lines
In the beginning of the specific analysis the gene expression values were scaled and the log2 fold changes were computed. For the downstream analysis the mean fold-change values across all cell lines treated with cisplatin were used. 

```{r}
log2FC.treated.untreated <- (treated.scaled-untreated.scaled)
```

The applied criterium to identify certain genes as biomarker for cisplatin is the expression fold change between tumor samples and treated tumor samples. The cisplatin treatment was the same for all cell lines: 15000nM and 24 hours.
The genes, which gene expression changes throughout all cell lines upon cisplatin pertubation the most also show the most extreme fold change values. So, genes were ranked after their fold-change and the 20 highest and 20 lowest were selected. 

`````{r, echo = F}
cisplatin.col=c()
j=1

while(j<820)
{
  if(isTRUE(meta[j,3]== "cisplatin"))
  {cisplatin.col= c(cisplatin.col,j)
  }
  j = j +1
}

FC.cisplatin = log2FC.treated.untreated[,cisplatin.col]

mean.FC.genes <- apply(FC.cisplatin, 1, mean)
mean.FC.celllines <- apply(FC.cisplatin, 2, mean)
```````

The resulting genes and their expression fold-changes between tumor samples and treated tumor samples were visualized through a barplot. 

````{r, echo = F, include = T, fig.width= 4.5, fig.height = 5, fig.show = 'hold'}
sd.cisplatin = apply(FC.cisplatin, 1, sd)
mean.FC.genes.sd = as.data.frame(cbind(mean.FC.genes, sd.cisplatin))
mean.FC.ordered = mean.FC.genes.sd[order(mean.FC.genes.sd$mean.FC.genes, decreasing=TRUE), ]

lowest.FC = mean.FC.ordered[13280:13299,]
par(mar = c(5, 7, 5, 5))
barplot(lowest.FC$mean.FC.genes,
        horiz = TRUE,
        xlim = c(-1.0, 0),
        main= "lowest log2 FC-values",
        xlab= "mean log2FC values in different celllines",
        names.arg = rownames(lowest.FC),
        col= "firebrick",
        las=1,
        border = "white", 
        cex.names =0.8)

highest.FC = mean.FC.ordered[1:20,]
par(mar = c(5, 10, 5, 5))
barplot(highest.FC$mean.FC.genes,
        horiz = TRUE,
        xlim = c(0, 1),
        main= "highest log2 FC-values",
        xlab= "mean log2FC values in different celllines",
        names.arg = rownames(highest.FC),
        col= "lightgreen",
        las=1,
        border = "white",
        cex.names =0.8)
``````

`````{r, echo= F}
#matrix containing the biomarker found through the FC 
highest.FC = as.matrix(highest.FC) 
lowest.FC = as.matrix(lowest.FC)
biomarker1.FC = as.matrix(rbind(highest.FC, lowest.FC))
highest.names <- row.names(highest.FC)
lowest.names <- row.names(lowest.FC)
row.names(biomarker1.FC) <- c(highest.names, lowest.names)
biomarker1 <- c(highest.names, lowest.names)
`````


### Additional criterium: Gene expression change in the same "direction"
In the second criterium gene expression changes unpon treatment were related to up- or downregulation. Consistent transcriptional changes can be defined as those in which gene expression of the majority of cell lines changed in the same direction, and only a small percentage of cell lines had a change in the opposite direction (Monks et. al.). 
So genes, which gene expression changes in 50 of 55 cell lines in the same direction (up or down regulation) were selected. 


```{r, echo=T}
is.neg = FC.cisplatin<0
i =1
j=1
a=1
biomarker2.up = c()
biomarker2.down = c()
while(j<13300){
  while(i<56){
    if(is.neg[j,i]==TRUE)
     {a=a+1}
    i=i+1}
  if(a>49)
     {biomarker2.down= c(biomarker2.down, j)}
  if(a<6)
    {biomarker2.up= c(biomarker2.up, j)}
  a=1
  i=1
  j=j+1}
```

````{r, echo =F}
biomarker2 = row.names(FC.cisplatin[c(biomarker2.down, biomarker2.up),])
`````


### Subselection of biomarker found through the FC-criterium 
The genes found through the FC-criterium (criterium 1) were subselected with our additional criterium, which narrows down the previously found genes (40) to 15 biomarker genes. With the second criterium the **keine ahnung wie man das schön formulieren soll**. 

Specifies the first criterium - gene expression changes in the same direction (cisplatin might work especially for patients with a gene expression in the same direction)

```{r, echo = F, include=TRUE}
i=1
j=1
a=1
double.biomarker = c()
while(i<41)
{
  while(j<375)
      {
        if(isTRUE(biomarker2[j] == biomarker1[i]))
           {double.biomarker = c(double.biomarker, biomarker1[a])
             }
            j = j +1
         }
          j= 1
          i=i+1
          a=a+1
}

double.biomarker.FC = FC.cisplatin[double.biomarker,]

```
`````{r, echo = F, include = T}
grid.newpage()
venn.plot <- draw.pairwise.venn(
  area1 = 40,
  area2 = 668,
  cross.area = 15,
  fill = c("light green", "light blue"),
  category = c("FC-Criterium", "Same Direction"),
  lty = "blank",
  cex = 2,
  cat.cex = 1.5,
  cat.pos = c(0, 1),
  cat.dist = -0.5,
  cat.just = list(c(0.5, 9), c(1,1.5)),
  cat.col = c("light green", "light blue"), 
  ext.pos = -3,
  ext.dist = -0.1,
  ext.length = 0
)
grid.draw(venn.plot)
```````

````{r, echo = T}
print(double.biomarker)
````

  


## Step 1b: How significant are the genes previously identified as biomarker? <a name="significance"></a> 

### Welch t-test to verify significance of the genes  

With a welch two sample t-test the significance of the expressional change of biomarker genes was verified. 
```{r, echo = F}
treated.cisplatin <- treated.scaled[,grep ("cisplatin", colnames(treated.scaled))]
untreated.cisplatin <- untreated.scaled[,grep("cisplatin", colnames(treated.scaled))]
```

**1. Checking normality**  
Bevore performing the t-test, normality was checked for all biomarkers through QQplots. This is an example for the POLR3B gene, which appears to be normally distributed.   
````{r, echo = F, include = T, dpi = 90, fig.align = 'center'}
par(mar = c(5, 5, 5, 5))
qqnorm(treated.cisplatin["POLR3B", ], main = "QQplot to check normality of the POLR3B gene")
qqline(treated.cisplatin["POLR3B", ])
legend(-2.5, 1.1,  cex = 0.9, box.lty = 0,
legend = "Gene expression values for POLR3B
in different celllines after cisplatin treatment.")
````

**2. Welch two sample t-test**  
Welch-t-test because the values are normally distributed, but the variances of the two samples are not equal and welch t-test corrects this problem by estimating the variances and adjusting the degrees of freedom used in the test. Paired two-sample-two-sided-t-test. 

````{r dpi =90, fig.align = 'center'}
pvalues.welch <- sapply(double.biomarker, function(x){
       t.test(treated.cisplatin[x,], untreated.cisplatin[x,],paired= T)$p.value
   })
plot(density(pvalues.welch), main = "Density of p-Values")
``````

The resulting p-values indicate significant expressional change of biomarkers in all celllines upon cisplatin pertubation. 


### Standard deviation to check consistency of fold-changes across cell lines    
Furthermore, the fold-change values for biomarker genes upon cisplatin pertubation were related to their standard deviation.  
`````{r, echo = F, include= F}
sd.cisplatin = apply(FC.cisplatin, 1, sd)
mean.FC.genes.sd = as.data.frame(cbind(mean.FC.genes, sd.cisplatin))
mean.FC.ordered = mean.FC.genes.sd[order(mean.FC.genes.sd$mean.FC.genes, decreasing=TRUE), ]
head(mean.FC.genes.sd)
plot(density(sd.cisplatin))
```

```{r, echo = T, include = T}
mean.double <- apply(double.biomarker.FC, 1, mean)
sd.double <- apply(double.biomarker.FC, 1, sd)
double = cbind(mean.double, sd.double)
colnames(double) = c("mean FC","SD" )
print(double)
`````

The resulting standard deviation values are higher than expected, which shows that so the fold-changes vary a lot between cell lines. 
A tissue based separation and selection of cancers, which are regularly treated with cisplatin (testicular carcinoma, ovarian and bladder cancer), might result in fold-change values with less variance and also in different and more specific biomarker genes. 

--> Different result as in T-test? (muss das noch rein?)

***

## Step 2: Influence of cisplation on the biomarkers gene expression in different cell lines <a name="upordown"></a> 
In the second step the influence of cisplatin treatment on up- or downregulation of the biomarkers gene expression was evaluated. The basis of this analysis were the log2-fold-changes. All cell lines were evaluated and related to tissue and cancer types. The result was visualized in a heatmap, which was clustered with kmeans-clustering. 

````{r, echo = F}
colfunc <- colorRampPalette(c("firebrick","firebrick3","lightcoral",
                              "lightyellow","lightskyblue1","steelblue1",
                              "steelblue3", "darkblue"))
`````

#### Checking how many clusters we will need in the heatmap (elbow plot) 
Previous to the heatmap computation, the optimal number of clusters was echecked with the help of an elbow plot. Unfortunately kmeans results in a high within-sum-of-square value, which indicates that the distance of data points in each cluster is high. (WIe könnte man sagen, dass das clustering nciht gut war?)

```{r, echo = T, include = T}
wss = sapply(2:7,function(k) {
  kmeans(x = t(double.biomarker.FC), centers = k)$tot.withinss
})
kmeans= kmeans(x = t(double.biomarker.FC), centers = 5, nstart = 10)
print(kmeans$tot.withinss)

`````

````{r, echo = F, include = T, dpi = 80, fig.align= 'center'}
plot(2:7, wss, type = "b",
     pch = 19,
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares",
     main = "Elbow plot for kmeans clustering of celllines")
````

#### Influence of cisplatin on the biomarkers gene expression
The following heatmap shows the influence of cisplatin on up- or downregulation of the biomarkers gene expression. The annotation bar relates cell lines to cancer types, but no patterns between tissues could be found. Genes can be clustered in two main groups: 
* 13 genes were mostly upregulated or showed a similar gene expression as in untreated samples  
  + LRBA: related to the immune checkpoint CTLA4, leads to a downregulation of immune surveillance  
  + CUX1: leads to a downregulation of e.g. the cell cycle  
* 2 genes were mostly downregulated upon cisplatin pertubation  
  + GMDS: GMDS mutation/deficiency is said to be important for the escape from NK cell mediated tumor surveillance  


As cisplatin acts as an intrastrand DNA-crosslinker, the targeted cells are not tumor-specific. Therefore this inconclusive result of gene up- and downregulation could be caused by the presence of more necrotic cells in the tumor microenvironment, which leads other tumor cells to downregulate immune surveillance upon cisplatin treatment. 

```{r dpi = 200, fig.align='center'}
colnames(double.biomarker.FC) <- meta[95:149,2]
annotation = data.frame(Cancertype = cellline$Cancer_type)
rownames(annotation) = cellline$Cell_Line_Name

pheatmap(double.biomarker.FC,
         color = colfunc(25),
         cluster_cols = TRUE,
         clustering_rows = TRUE,
         clustering_method ="ward.D2",
         treeheight_row = 30,
         treeheight_col = 30,
         annotation_col = annotation,
         legend = T,
         legend_breaks = c(-1,1),
         legend_labels = c("down", "up"),
         show_colnames = F,
         cutree_rows = 2,
         cutree_cols = 4,
         border_color = "white",
         scale = "column")
```````


***

## Step 3: Further analysis of biomarker <a name="alterations"></a> 

Is there an amplification, deletation or neutral gene copy number? 
Values smaller than -1 indicate a deletation, values higher than 1 an amplification. All values in between -1 and 1 represent a neutral gene copy number as the values are already log2... QUELLE. 


````{r}
copynumber.biomarker = as.matrix(copynumber[double.biomarker,])
copynumber.quali = ifelse(copynumber.biomarker <= (-1), (-1), ifelse (copynumber.biomarker >= (1), 1, 0))
````

### Distribution of genetic alterations in cisplatin biomarker genes
Visualizing the distribution of genetic alterations. 
````{r, echo = F, include = T, dpi = 100}
quantiles = as.matrix(quantile(copynumber.biomarker))
d <- density(copynumber.biomarker)
plot(d, main = "Distributions of genetic alterations in biomarker genes")
abline(v= c(-1,1), col= c("red", "blue"), lty =2)
legend(-6, 1.2, legend = c("Deletion", "Amplification"), col = c("red", "blue"), lty = 2)
`````

### checking the optimal number of cluster (elbow plot) 
kmeans clustering of the heatmap.. 
````{r, echo = F, include = T, warning = F, dpi = 80}
wss = sapply(2:7, function(k) {
  kmeans(x = t(copynumber.quali), centers = k)$tot.withinss
})
plot(2:7, wss, 
     type = "b", pch = 19, 
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares",
     main = "Elbowplot kmeans clustering - biomarker gene alterations")

kmeans= kmeans(x = t(copynumber.quali), centers = 5, nstart = 10)
kmeans$tot.withinss
`````


  
  
### Connection between biomarker and genetic alterations
--> Geninterpretation
--> Gentherapie würde was anderes treffen

````{r fig.align = 'right', echo = F, include = T, dpi = 160}
colfunc2 <- colorRampPalette(c("firebrick2", "grey88", "deepskyblue3"))

colnames(double.biomarker.FC) <- meta[95:149,2]
annotation = data.frame(Cancertype = cellline$Cancer_type)
rownames(annotation) = cellline$Cell_Line_Name

pheatmap(copynumber.quali,
         color = colfunc2(3),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         clustering_method = "ward.D2",
         legend = TRUE,
         legend_breaks = c(-1,0,1),
         legend_labels = c("deletion","", "amplification"),
         border_color = "white",
         show_colnames = FALSE,
         cutree_rows = 4,
         cutree_cols = 3,
         treeheight_row = 20,
         treeheight_col = 20,
         annotation = annotation)
````
</div>
## Sources

"The NCI Transcriptional Pharmacodynamics Workbench: A Tool to Examine Dynamic Expression Profiling of Therapeutic Response in the NCI-60 Cell Line Panel.", Monks et. al.